---
title: "Compare Parcel Pop Methods"
format: gfm
editor: visual
---

## A 'real' data example

Our real data examples will be examining the percent of deaths between 2022 and 2025 that were due to diabetes and the count of deaths due to all drugs during that period. The source geography will be tracts and the target geography will be HRAs. Records that are not geocoded within King County are omitted from the analysis.

```{r, include=FALSE}
library('DBI')
library('sf')
library('apde.data')
library('data.table')
library('dplyr')
library('ggplot2')
library('spatagg')

tract = st_read("//dphcifs/APDE-CDIP/Shapefiles/Census_2020/tract/kc_tract.shp", quiet = T)
hra = st_read("//dphcifs/APDE-CDIP/Shapefiles/HRA/hra_2020.shp", quiet = T)

# Load death data
con <- authenticate_hhsaw()
    
deaths <- setDT(
  DBI::dbGetQuery(
    con,
    "
    select fa.[state_file_number], chi_year, diabetes, 
    all_drugs, res_lat as lat, res_long as long
    from death.final_analytic as fa

    where chi_year >=2022 AND chi_year<=2025 and
    res_lat is not null and res_long is not null
    and res_geo_source IS NOT NULL
    and res_geo_source NOT IN ('0', '5 DIGIT URBAN ZIP CE', 'CITY CENTROID', 
    'CITY OR PLACE CENTRO', 'CNTY EXCEP', 'POST OFFICE CENTROID', 
    'STREET CENTROID', 'ZIP CENTROID', 'ZIP PLUS 4 CENTROID',
    'ZIP4 CENTROID', 'Null') AND
    chi_geo_kc = 'King County'

    
    
    "
  )
)

deaths = st_as_sf(deaths, coords = c('long', 'lat'), crs = 4326)
kc = st_read('//Kcitfsrprpgdw01/kclib/Plibrary2/politicl/shapes/polygon/kingco.shp', quiet = T)
kc = st_transform(kc, st_crs(hra))
deaths = st_transform(deaths, st_crs(hra))
deaths = st_intersection(deaths, summarize(kc))

PN = function(x) prettyNum(x, big.mark = ',')
```

## The data

In the extract, there are `r PN(nrow(deaths))` deaths from all causes, `r PN(sum(deaths$diabetes ==1, na.rm = T))` from diabetes and `r PN(sum(deaths$all_drugs ==1, na.rm = T))` from all drugs.

```{r, echo = FALSE}

deaths = deaths %>% mutate(cause = case_when(diabetes ==1 ~ 'Diabetes', all_drugs ==1 ~ 'Drugs', TRUE ~ 'Other'))

g = ggplot() + 
  geom_sf(data = kc) + 
  geom_sf(data = deaths, aes(color = cause)) +
  scale_color_brewer(type = 'qual', palette = 2) +
  theme_bw() +
  ggtitle('Deaths in KC, 2022 - 2025')

g
```

## The truth

```{r}
# Tract level
tract = st_transform(tract, st_crs(deaths))
dtract = st_join(deaths, tract[,'GEOID'])
dtract = dtract %>% 
  st_drop_geometry() %>% 
  group_by(GEOID) %>%
  summarize(diabetes = sum(diabetes,na.rm =T),
            drugs = sum(all_drugs, na.rm = T),
            all = n())

dtract = dtract %>% mutate(pdiabetes := diabetes/all)

# HRA

dhra = st_join(deaths, hra[,'id']) %>%
  st_drop_geometry() %>%
  group_by(id) %>%
   summarize(diabetes = sum(diabetes,na.rm =T),
            drugs = sum(all_drugs, na.rm = T),
            all = n())%>% 
  mutate(pdiabetes := diabetes/all) %>%
  filter(!is.na(id))
```

## Crosswalks

```{r}

new_pp = read_sf("//dphcifs/APDE-CDIP/Shapefiles/historical_parcel/new_parcel_pop.gpkg")
old_pp = read_sf("//dphcifs/APDE-CDIP/Shapefiles/historical_parcel/old_parcel_pop.gpkg")

geog = create_xwalk(tract, hra, 'GEOID', 'id', method = 'fractional overlap')
ppop_new = create_xwalk(tract, hra, 'GEOID', 'id', method = 'point pop', point_pop = new_pp) |>
  filter(isect_amount >0)
ppop_old = create_xwalk(tract, hra, 'GEOID', 'id', method = 'point pop', point_pop = old_pp) |>
  filter(isect_amount >0)


# go from tract to hra
t2hra = function(dtract, xw,  type = 'geog'){
  
  r = lapply(c('diabetes', 'drugs', 'all', 'pdiabetes'), function(x){
    
    d = crosswalk(dtract, 'GEOID', est = x, 
              proportion = x == 'pdiabetes', 
              xwalk_df = xw)
    
    d$var = x
    
    d
    
  })
  
  r = rbindlist(r)
  r = dcast(r, target_id ~ var, value.var = 'est')
  r[, 'type' := type]
  
  r
}

rgeog = t2hra(dtract, geog, 'geog')
rppop_new = t2hra(dtract, ppop_new, 'ppop_new')
rppop_old = t2hra(dtract, ppop_old, 'ppop_old')

rmse = function(obs, pred) round(sqrt(mean((obs - pred)^2)),3)
# compute rmse for stuff
compare_to_truth = function(obs, pred){
  pred = copy(pred)
  setnames(pred, 
           c('diabetes', 'drugs', 'all', 'pdiabetes'),
           c('e_diabetes', 'e_drugs', 'e_all', 'e_pdiabetes')
           )
  
  r = merge(obs, pred, all.x = T, by.x = 'id', by.y = 'target_id')
  setDT(r)
  
  # convert two counts, and then compute
  r[, e_pdiabetes2 := e_diabetes/e_all]
  
  # compute RMSEs
  e = r[, .(all = rmse(all, e_all),
        diabetes = rmse(diabetes, e_diabetes),
        drugs = rmse(drugs, e_drugs),
        pdiabetes = rmse(pdiabetes, e_pdiabetes),
        pdiabetes2 = rmse(pdiabetes, e_pdiabetes2)), type]
  e
  
}

cgeog = compare_to_truth(dhra, rgeog)
cppop_new = compare_to_truth(dhra, rppop_new)
cppop_old = compare_to_truth(dhra, rppop_old)

knitr::kable(rbind(cgeog, cppop_old, cppop_new), label = 'RMSE by variable and approach')
```

As the table shows, parcel population has lower overall error across all indicators.

```{r}

dhra = merge(
  dhra, 
  rgeog[, .(id = target_id, geog_all = all)], all.x = T,
  by = 'id')

dhra = merge(
  dhra,
  rppop_new[, .(id = target_id, ppop_new_all = all)],
  all.x = T,
  by = 'id')

dhra = merge(
  dhra,
  rppop_old[, .(id = target_id, ppop_old_all = all)],
  all.x = T,
  by = 'id')

setDT(dhra)
ghra = melt(dhra[, .(id, all, geog_all, ppop_new_all, ppop_old_all)], id.vars = c('id', 'all'))

ghra[, dif := all - value]
ghra[, variable := factor(variable, c('geog_all', 'ppop_new_all', 'ppop_old_all'), c('Geographic', 'Parcel Pop (New)', 'Parcel Pop (Old)'))]
ghra = merge(ghra, hra[, 'id'], by = 'id')
ghra = st_as_sf(ghra)

ggplot(ghra, aes(fill = dif)) + 
  geom_sf() + 
  facet_wrap(~variable) +
  scale_fill_viridis_c() +
  theme_bw() +
  ggtitle('Obs - Pred by crosswalk type', 'All deaths') +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

This graph shows the difference in the observed number of deaths relative to the "predicted" number of deaths created via the crosswalking approaches. In general, the parcel/point population approach has less "error" than the geographic approach.
